---
title: "Base R"
subtitle: "An introduction"
author: "Kevin Rue-Albrecht"
institute: "University of Oxford"
date: "2020/02/03 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Data types and structures

R has 6 basic data types.

- character

- numeric (real or decimal)

- integer

- logical

- complex

- raw

---

# The character type

```{r}
"text"
"a longer 'character' value"
'Single quotes can be used to escape "double quote"'
```

---

# The numeric type

The numeric type can represent values with decimal point.

```{r}
1.05
```

Integers can also be stored using as numeric values.

```{r}
3.000
```

---

# The integer type

The `L` tells R to store this as an integer.

```{r}
3L
is.integer(3L)
is.integer(3)
```

A numeric value stored as an integer loses digits after the decimal point.

```{r}
as.integer(3.5)
```

---

# The logical type

```{r}
TRUE
FALSE
```

---

# The complex type

```{r}
1+3i
```

---

# The raw type

The raw type is intended to hold raw bytes.

A raw vector is printed with each byte separately represented as a pair of hexadecimal digits (used to simplify how binary is represented).

```{r}
as.raw(40)
charToRaw("A")
```

```{r}
charToRaw("A test string")
```

---

# Data frames

Data frames represent collections of variables in tabular format.

Each row of a data frame generally represents an observation, while each column describes a different aspect of that observation.

A number of toy data sets are provided with the R installation.

```{r}
library(datasets)
data(iris)
head(iris)
```

